A real-time particle filter for modeling travel times of transit vehicles through a GTFS-based road network

### 2018-11-01 anzjs2018/sections/04_implementation.tex
run spellcheck

The general structure of the \rt component is shown below,
with the bold steps being those discussed in this paper.
\begin{enumerate}
\item Load GTFS data from database
\item Each time new data are recieved \ldots
\begin{enumerate}
    \item Update or create new vehicle objects from the new data
    \item \textbf{Run particle filter on each vehicle to update or initialise state}
    \item \textbf{Update state of any roads for which vehicles 
        have completed travel}
    \item Generate ETAs for vehicles uing combination of particle filter and network state
    \item Write ETAs to extended Google Protobuf binary file for distribution
\end{enumerate}
\end{enumerate}

During the development of the application,
we were primarly concerned with ensuring each component of the program
is as efficient as possible,
allowing ETAs to be generated and distributed fast enough to be faesible in real-time,
with a target of 30~seconds or faster at peak time.
Using C++ provides the memory management control necessary to make the application
viable in real-time.
We also make use of OpenMP for parallisation,
since each vehicle is modelled independently so it is trivial to scale up without
worrying about threadsafety.

@@

In each iteration, 
the timing of the various program components is recorded.
Since the number of vehicles traveling at any given time changes throughout the day,
we used the average timings over an off-peak 15~minute window.
Figure~\ref{fig:timings} shows the average timings for 
varying numbers of particles, $N$.

@@

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/04_model_results_dist.pdf}
    \caption{
        The distance between the observation and the nearest point on the shape
        is calculated by finding the shortest distance between point and path.
        This gives an approximate idea of GPS error, however it should be noted
        that many observations are made automatically at waypoints,
        which accounts for the peak at 1m.
        Note also that these observations have been truncated at 20m.
    }
    \label{fig:dist_to_route}
\end{figure}

From Figure~\ref{fig:degen_rate}, we see that $N_\text{eff}$ increases
with GPS error and decreases with system noise.
The degeneration rate decreases with GPS error and increases with $N$,
but is relatively unaffected by system noise.
Large GPS error affects the likelihood,
and gives more weight to particles farther from the observation
than does a smaller GPS error.
Conversesly, increasing system noise spreads out the particle cloud,
so fewer particles will be near the vehicle.
Thus, these results are not surprising, 
but show that the model is working as expected.

@@

The last measure of performance is the relative variance of segment travel times.
This is calculated by computing the average variance of travel times over all simulations,
and then for each simulation, 
taking the average (over road segments) of the ratio between variance for that simulations
and variance for all simualtions.
Figure~\ref{fig:travel_times} shows that simulations with larger GPS error
result in a higher relative variance,
while simulations with more particles have lower relative variance.
There is no visible effect of system noise on travel time variation.

>>>>

The general structure of the \rt component is shown below,
with the bold steps being those discussed in this paper.
\begin{enumerate}
\item Load GTFS data from database
\item Each time new data are received \ldots
\begin{enumerate}
    \item Update or create new vehicle objects from the new data
    \item \textbf{Run particle filter on each vehicle to update or initialise state}
    \item \textbf{Update state of any roads for which vehicles 
        have completed travel}
    \item Generate ETAs for vehicles using combination of particle filter and network state
    \item Write ETAs to extended Google Protobuf binary file for distribution
\end{enumerate}
\end{enumerate}

During the development of the application,
we were primarily concerned with ensuring each component of the program
is as efficient as possible,
allowing ETAs to be generated and distributed fast enough to be feasible in real-time,
with a target of 30~seconds or faster at peak time.
Using C++ provides the memory management control necessary to make the application
viable in real-time.
We also make use of OpenMP for parallelisation,
since each vehicle is modelled independently so it is trivial to scale up without
worrying about thread safety.

@@

In each iteration, 
the timing of the various program components is recorded.
Since the number of vehicles travelling at any given time changes throughout the day,
we used the average timings over an off-peak 15~minute window.
Figure~\ref{fig:timings} shows the average timings for 
varying numbers of particles, $N$.

@@

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/04_model_results_dist.pdf}
    \caption{
        The distance between the observation and the nearest point on the shape
        is calculated by finding the shortest distance between point and path.
        This gives an approximate idea of GPS error, however it should be noted
        that many observations are made automatically at way points,
        which accounts for the peak at 1m.
        Note also that these observations have been truncated at 20m.
    }
    \label{fig:dist_to_route}
\end{figure}

From Figure~\ref{fig:degen_rate}, we see that $N_\text{eff}$ increases
with GPS error and decreases with system noise.
The degeneration rate decreases with GPS error and increases with $N$,
but is relatively unaffected by system noise.
Large GPS error affects the likelihood,
and gives more weight to particles farther from the observation
than does a smaller GPS error.
Conversely, increasing system noise spreads out the particle cloud,
so fewer particles will be near the vehicle.
Thus, these results are not surprising, 
but show that the model is working as expected.

@@

The last measure of performance is the relative variance of segment travel times.
This is calculated by computing the average variance of travel times over all simulations,
and then for each simulation, 
taking the average (over road segments) of the ratio between variance for that simulations
and variance for all simulations.
Figure~\ref{fig:travel_times} shows that simulations with larger GPS error
result in a higher relative variance,
while simulations with more particles have lower relative variance.
There is no visible effect of system noise on travel time variation.

### 2018-11-01 anzjs2018/sections/06_discuss.tex
run spellcheck

In this paper we have described an approach to transit vehicle modeling
that enables the \rt estimation of road state from vehicle position data,
which can later be used to make arrival time predictions
that account for \rt traffic conditions.
The vehicle model, implemented using a particle filter,
allows complex bus behaviours to be modeled,
with future work planned to extend the model to handle physical intersections,
such as traffic lights.
From the particle filter,
vehicle travel times along roads are estimated which are in turn used 
in the modeling of the road network state.

The network model presented here is simple,
but provides the foundation for future work to
devel a formal model combining \rt travel time information with historical data, 
allowing short- and medium-term forecasts of road state,
for example incorporating peak time congestion.
This in turn will provide the basis for an arrival time prediction model
that combines \rt vehicle state with travel time forecasts
and provide commuters with reliable ETAs.
These will likely consist of a point estimate along with a prediction interval,
as it is impossible given the available data to accurately predict arrival time
due to traffic lights, unknown dwell times at intermediate bus stops,
and many other factors.

% The main goal of this work was to explore the faesibility
% of a particle filter as a \rt estimation method for transit vehicle states,
% and the use of estimated states to determine traffic conditions
% throughout the network.
% This allows vehicles to inform others---irrespective of route---of 
% current \rt traffic conditions,
% and provide a means of obtaining more reliable ETAs.

@@

% The network model is still in the early stages of development,
% but given that the implementation is running well within our 30~second target,
% there is room for developing a more complex model.
% Future work will explore the use of historical data to develop informative priors
% for travel times,
% which will provide a useful fallback in the case of no data,
% or for making more accurate long-term predictions.
% Once such a model has been developed, we plan to explore prediction estimates,
% particuarly comparing point and intervals as a means of communicating 
% uncertainty to commuters.

>>>>

In this paper we have described an approach to transit vehicle modelling
that enables the \rt estimation of road state from vehicle position data,
which can later be used to make arrival time predictions
that account for \rt traffic conditions.
The vehicle model, implemented using a particle filter,
allows complex bus behaviours to be modelled,
with future work planned to extend the model to handle physical intersections,
such as traffic lights.
From the particle filter,
vehicle travel times along roads are estimated which are in turn used 
in the modelling of the road network state.

The network model presented here is simple,
but provides the foundation for future work to
develop a formal model combining \rt travel time information with historical data, 
allowing short- and medium-term forecasts of road state,
for example incorporating peak time congestion.
This in turn will provide the basis for an arrival time prediction model
that combines \rt vehicle state with travel time forecasts
and provide commuters with reliable ETAs.
These will likely consist of a point estimate along with a prediction interval,
as it is impossible given the available data to accurately predict arrival time
due to traffic lights, unknown dwell times at intermediate bus stops,
and many other factors.

% The main goal of this work was to explore the feasibility
% of a particle filter as a \rt estimation method for transit vehicle states,
% and the use of estimated states to determine traffic conditions
% throughout the network.
% This allows vehicles to inform others---irrespective of route---of 
% current \rt traffic conditions,
% and provide a means of obtaining more reliable ETAs.

@@

% The network model is still in the early stages of development,
% but given that the implementation is running well within our 30~second target,
% there is room for developing a more complex model.
% Future work will explore the use of historical data to develop informative priors
% for travel times,
% which will provide a useful fallback in the case of no data,
% or for making more accurate long-term predictions.
% Once such a model has been developed, we plan to explore prediction estimates,
% particularly comparing point and intervals as a means of communicating 
% uncertainty to commuters.
